Black swans in ecological time series\\ Black swans in population dynamics\\ Heavy-tailed ecological time series\\ Black swans in ecological dynamics\\ Black swan ecological dynamics\\ \ldots

### 2014-10-30 text/ms.tex
Fix line number - align conflict in preamble
Minor edits to ms

\noindent
\emph{Main message, to be deleted (25 words)}: Black swans are present \&
taxonomically widespread but rare in population dynamics. Extreme climate,
predation, parasites, and their interactions are common causes; intrinsic
drivers are unclear.

@@

Black swans are statistically improbable events that nonetheless occur ---
often with profound implications. While extremes in the physical environment,
such as monsoons and heat waves, are widely studied and increasing in magnitude
and frequency, it remains unclear the extent to which ecological populations
buffer or suffer from extremes. Here, we develop a probability model to
estimate the degree of heavy-tailedness (presence of black swans) in ecological
process noise. We apply our model to \NPops~time series from around the world
across \NOrders~taxonomic orders and seven classes. We find strong evidence of
black swans, but they are rare, occurring in
\overallMinPerc--\overallMaxPerc\%of populations; most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects
(\InsectaRangePerc\%), and fishes(\OsteichthyesRangePerc\%). When they occur,
they tend to be driven by climate and severe winters, cycles of parasites and
predators, and interactions between these elements. They are more frequently
detected for populations with longer time series and lower levels of process
noise. We find little evidence of intrinsic life-history covariates. TODO close
with modelling and policy implications

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
represents population growth as a linear function in $\log$ space. If we let
$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
Gompertz model as:

\vspace{-6mm}
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
\end{align*}
The parameter $\lambda$ represents the expected $\log$ number of new
individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
$b = 1$, maximally density dependent if $b = 0$, and inversely density
dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
the Gompertz has proved to be the most frequently parsimonious population model
\citep{brook2006}.

We extended the Gompertz model by allowing the process noise to be
drawn from a Student-t distribution with degrees of freedom parameter $\nu$,
and scale parameter $\sigma$:

\vspace{-6mm}
\begin{align*}
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
If $\nu$ is small ($\lesssim 10$), the t distribution has much heavier tails
than a normal distribution. For example, at $\nu = 2$, the probability of
drawing a value more than five standard deviations below the mean is $1.8$\%,
whereas the probability of drawing such a value from a normal distribution is
nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the
distribution approaches the normal distribution. By estimating the value of
$\nu$ we can quantity how heavy-tailed the process noise deviations are
(Fig.~\ref{fig:didactic}).

@@

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines focussed on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans (e.g.~REF-TODO). A similar focus might benefit resource management once our
ability to predict the spatial probability and covariates of ecological black
swans improves. Second, we can choose management policy that is robust to heavy
tails. For example, setting target population abundances that are appropriately
far from critical limits may buffer black swan events
\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
diversity may allow some components of populations persist when others are
affected by disease or extreme environmental forces
\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and unexpected,
surprising, or counterintuitive ecological dynamics offer a tremendous
opportunity to learn about ecological systems, evaluate when our models break
down, and adjust future management policy \citep{doak2008, pine-iii2009,
  lindenmayer2010}. In recent decades ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggests that an added focus on extremes
represents the next frontier, particularly in the face expected increases in
climate extremes \citep{meehl2004,ipcc2012}.

>>>>

%\noindent
%\emph{Main message, to be deleted (25 words)}: Black swans are present \&
%taxonomically widespread but rare in population dynamics. Extreme climate,
%predation, parasites, and their interactions are common causes; intrinsic
%drivers are unclear.

@@

Black swans are statistically improbable events that nonetheless occur ---
often with profound implications. While extremes in the physical environment,
such as monsoons and heat waves, are widely studied and increasing in magnitude
and frequency, it remains unclear the extent to which ecological populations
buffer or suffer from extremes. Here, we develop a probability model to
estimate the degree of heavy-tailedness (presence of black swans) in ecological
process noise. We apply our model to \NPops~time series from around the world
across \NOrders~taxonomic orders and seven classes. We find strong evidence of
black swans, but they are rare, occurring in
\overallMinPerc--\overallMaxPerc\%of populations; most frequently for birds
(\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects
(\InsectaRangePerc\%), and fishes(\OsteichthyesRangePerc\%). When they occur,
they tend to be driven by climate and severe winters, cycles of parasites and
predators, and interactions between these elements. They are more frequently
detected for populations with longer time series and lower levels of process
noise. We find little evidence of intrinsic life-history covariates. \textit{TODO close
with modelling and policy implications. Something like:
The presence of black swans highlights the importance of developing management strategies that detect quickly, respond to, and are robust to extremes in population dynamics --- particularly as the frequency and magnitude of climatic extremes increase over the next century.}

@@

Our main analysis focuses on the commonly applied Gompertz population dynamics
model \citep[e.g.][]{knape2012,dennis2014,connors2014}. The Gompertz model
represents population growth as a linear function in $\log$ space. If we let
$x_t$ represent the $\log$ abundance ($N$) at time $t$, we can represent the
Gompertz model as:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Normal}(0, \sigma^2).
\end{align*}
The parameter $\lambda$ represents the expected $\log$ number of new
individuals at time $t + 1$ if $N_t = 1$. The model is density independent if
$b = 1$, maximally density dependent if $b = 0$, and inversely density
dependent if $b < 0$. The process noise $\epsilon_t$ is modelled as normally
distributed with standard deviation $\sigma$. In previous analyses of the GPDD,
the Gompertz has proved to be the most frequently parsimonious population model
\citep{brook2006}.

We extended the Gompertz model by allowing the process noise to be
drawn from a Student-t distribution with degrees of freedom parameter $\nu$,
and scale parameter $\sigma$:
\begin{align*}
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma).
\end{align*}
If $\nu$ is small ($\lesssim 10$), the t distribution has much heavier tails
than a normal distribution. For example, at $\nu = 2$, the probability of
drawing a value more than five standard deviations below the mean is $1.8$\%,
whereas the probability of drawing such a value from a normal distribution is
nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the
distribution approaches the normal distribution. By estimating the value of
$\nu$ we can quantity how heavy-tailed the process noise deviations are
(Fig.~\ref{fig:didactic}).

@@

Recognizing the prevalence of heavy-tailed dynamics suggests a number of policy
directions. First, ecological resource management can draw from other
disciplines focussed on heavy tails. For example, earthquake preparedness and
response is focussed on black swan events. Similarly to ecological black swans,
we can rarely predict the specific location and timing of large earthquakes.
But, earthquake preparedness involves spatial planning based on forecast
probabilities to focus early detection efforts and develop disaster response
plans (e.g.~REF-TODO). A similar focus might benefit resource management once our
ability to predict the spatial probability and covariates of ecological black
swans improves. Second, we can choose management policy that is robust to heavy
tails. For example, setting target population abundances that are appropriately
far from critical limits may buffer black swan events
\citep[e.g.][]{caddy1996}, and maintaining genetic, phenotypic, and behavioural
diversity may allow some components of populations persist when others are
affected by disease or extreme environmental forces
\citep[e.g.][]{schindler2010, anderson2014}. Finally, extreme and unexpected,
surprising, or counterintuitive ecological dynamics offer a tremendous
opportunity to learn about ecological systems, evaluate when our models break
down, and adjust future management policy \citep{doak2008, pine-iii2009,
  lindenmayer2010}. In recent decades ecology has moved toward focussing on
aspects of variance in addition to mean responses \citep[e.g.][]{loreau2010a,
  thompson2013}. Our results suggest that an added focus on extremes
represents the next frontier, particularly in the face expected increases in
climate extremes \citep{meehl2004,ipcc2012}.
